---
title: "DATA 621 - Final Project"

author: "Peter"
output: 
  html_document:
    theme: cerulean
    highlight: pygments
    css: ./lab.css
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, 
                      message = F,
                      echo = F,
                      fig.align = "center")
```

# Introduction

For this project I will be using Student performance dataset located at UCI Machine Learning Repository. The repository has 2 datasets one for Mathematics `student-mat.csv` and the other one for Portuguese language `student-por.csv`. In this project we will use these datasets and create models to predict the grades in mathematics and portuguese based on the student grades, demographic, social and school related features.

# Data exploration

Attributes for both `student-mat.csv` (Math course) and `student-por.csv` (Portuguese language course) datasets:

1 `school` - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)

2 `sex` - student's sex (binary: 'F' - female or 'M' - male)

3 `age` - student's age (numeric: from 15 to 22)

4 `address` - student's home address type (binary: 'U' - urban or 'R' - rural)

5 `famsize` - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)

6 `Pstatus` - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)

7 `Medu` - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education)

8 `Fedu` - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education)

9 `Mjob` - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')

10 `Fjob` - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')

11 `reason` - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')

12 `guardian` - student's guardian (nominal: 'mother', 'father' or 'other')

13 `traveltime` - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)

14 `studytime` - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)

15 `failures` - number of past class failures (numeric: n if 1<=n<3, else 4)

16 `schoolsup` - extra educational support (binary: yes or no)

17 `famsup` - family educational support (binary: yes or no)

18 `paid` - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)

19 `activities` - extra-curricular activities (binary: yes or no)

20 `nursery` - attended nursery school (binary: yes or no)

21 `higher` - wants to take higher education (binary: yes or no)

22 `internet` - Internet access at home (binary: yes or no)

23 `romantic` - with a romantic relationship (binary: yes or no)

24 `famrel` - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)

25 `freetime` - free time after school (numeric: from 1 - very low to 5 - very high)

26 `goout` - going out with friends (numeric: from 1 - very low to 5 - very high)

27 `Dalc` - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)

28 `Walc` - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)

29 `health` - current health status (numeric: from 1 - very bad to 5 - very good)

30 `absences` - number of school absences (numeric: from 0 to 93)

These grades are related with the course subject, Math or Portuguese:

31 `G1` - first period grade (numeric: from 0 to 20)

32 `G2` - second period grade (numeric: from 0 to 20)

33 `G3` - final grade (numeric: from 0 to 20, output target)


```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(psych)
library(corrplot)
library(RColorBrewer)
library(knitr)
library(MASS)
library(caret)
library(kableExtra)
library(ResourceSelection)
library(pROC)
library(ggplot2)
library(gridExtra)
library(MASS)
library(scales)
library(rpart.plot)
library(ggplot2)
library(ggfortify)
library(gridExtra)
library(forecast)
library(fpp2)
library(fma)
library(kableExtra)
library(e1071)
library(mlbench)
library(ggcorrplot)
library(DataExplorer)
library(timeDate)
library(caret)
library(GGally)
library(corrplot)
library(RColorBrewer)
library(tibble)
library(tidyr)
library(tidyverse)
library(tidyselect)
library(dplyr)
library(reshape2)
library(mixtools)
library(tidymodels)
library(ggpmisc)
library(regclass)
library(pROC)
library(skimr)
library(naniar)
library(RANN)
library(kableExtra)
library(geoR)
library(pscl)
```




```{r}
maths = read.table("https://raw.githubusercontent.com/petferns/DATA621/main/student-mat.csv",sep=";",header=TRUE)
portuguese = read.table("https://raw.githubusercontent.com/petferns/DATA621/main/student-por.csv",sep=";",header=TRUE)
```

Dimension for Maths dataset

```{r}
dim(maths)
```
Dimension for Portuguese dataset

```{r}
dim(portuguese)
```

```{r}
summary(maths)
```

```{r}
summary(portuguese)
```

We do the cleaning of the datasets by replacing the yes/ no with 1/ 0 in both of the datasets. `maths_clean` and `port_clean`

```{r}
maths_clean <- maths

#Replace Yes/ no with 1/0
maths_clean$schoolsup <- ifelse(maths_clean$schoolsup=="yes", 1, 0)
maths_clean$famsup <- ifelse(maths_clean$famsup=="yes", 1, 0)
maths_clean$paid <- ifelse(maths_clean$paid=="yes", 1, 0)
maths_clean$activities <- ifelse(maths_clean$activities=="yes", 1, 0)
maths_clean$nursery <- ifelse(maths_clean$nursery=="yes", 1, 0)
maths_clean$higher <- ifelse(maths_clean$higher=="yes", 1, 0)
maths_clean$internet <- ifelse(maths_clean$internet=="yes", 1, 0)
maths_clean$romantic <- ifelse(maths_clean$romantic=="yes", 1, 0)

#Portuguese dataset

port_clean <- maths

#Replace Yes/ no with 1/0
port_clean$schoolsup <- ifelse(port_clean$schoolsup=="yes", 1, 0)
port_clean$famsup <- ifelse(port_clean$famsup=="yes", 1, 0)
port_clean$paid <- ifelse(port_clean$paid=="yes", 1, 0)
port_clean$activities <- ifelse(port_clean$activities=="yes", 1, 0)
port_clean$nursery <- ifelse(port_clean$nursery=="yes", 1, 0)
port_clean$higher <- ifelse(port_clean$higher=="yes", 1, 0)
port_clean$internet <- ifelse(port_clean$internet=="yes", 1, 0)
port_clean$romantic <- ifelse(port_clean$romantic=="yes", 1, 0)



```

# Data Visualization

## Correlation matrix

From the coorelation matrix for both of the datasets we see there is a strong positive relationship among the grades G1,G2 and G3. 
Variables- age, medu, fedu, traveltime,studytime,failures,schoolsup,famsup,paid, activities, nursery, higher, internet, romantic, famrel, freetime, goout, dalc, walc, health, absences are positively correlated and other are negatively correlated.

### Mathematics

```{r , fig.height = 7, fig.width = 7, echo=FALSE, warning=FALSE}
DataExplorer::plot_correlation(data = maths_clean,type = "all",cor_args = list("use" = "pairwise.complete.obs"))
```

### Portuguese


```{r , fig.height = 7, fig.width = 7, echo=FALSE, warning=FALSE}
DataExplorer::plot_correlation(data = port_clean,type = "all",cor_args = list("use" = "pairwise.complete.obs"))
```


## Histogram


From the below histograms for Maths and Portuguese dataset - we see grades G1, G2 and G3 are almost normally distributed. We see right skewing in `absences` in both of the dataset.


### Mathematics

```{r, fig.height = 7, fig.width = 7, echo=FALSE, warning=FALSE}
DataExplorer::plot_histogram(
  geom_histogram_args = list(alpha = 0.5),
   data = maths_clean,
         ggtheme=theme_bw())
```

### Portuguese

```{r, fig.height = 7, fig.width = 7, echo=FALSE, warning=FALSE}
DataExplorer::plot_histogram(
  geom_histogram_args = list(alpha = 0.5),
   data = port_clean,
         ggtheme=theme_bw())

```

## Boxplot

In the boxplot's below we see high intequartile range in variables - `activities`, `famsup`, `paid`, `romantic`

### Mathematics

```{r, fig.height = 7, fig.width = 7, echo=FALSE}
distribution <- select_if(maths_clean,is.numeric) %>% 
  gather(key = 'variable', value = 'value')
ggplot(distribution, aes(variable, value)) + 
  geom_boxplot() + 
  facet_wrap(. ~variable, scales='free', ncol=6)
```


### Portuguese

```{r, fig.height = 7, fig.width = 7, echo=FALSE}
distribution <- select_if(port_clean,is.numeric) %>% 
  gather(key = 'variable', value = 'value')
ggplot(distribution, aes(variable, value)) + 
  geom_boxplot() + 
  facet_wrap(. ~variable, scales='free', ncol=6)
```

# Model building

We split the cleaned datasets `maths_clean` and `port_clean` into a training set and a testing set, 80% will be set for training and 20% for testing set.

```{r}
#split the data with 80% for training and 20% for testing

set.seed(3)
maths_split <- initial_split(maths_clean, prop = 0.8)
maths_train <- training(maths_split)
maths_test <- testing(maths_split)

set.seed(3)
port_split <- initial_split(port_clean, prop = 0.8)
port_train <- training(port_split)
port_test <- testing(port_split)

```


```{r}
#common function to display residual plots 
residualPlots <- function(model) {
  
  if (is.null(model)) {
    return
  }
  
  layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))
  plot(residuals(model))
  hist(model[["residuals"]], freq = FALSE, breaks = "fd", main = "Residual Histogram",
       xlab = "Residuals",col="lightgreen")
  lines(density(model[["residuals"]], kernel = "ep"),col="blue", lwd=3)
  curve(dnorm(x,mean=mean(model[["residuals"]]), sd=sd(model[["residuals"]])), col="red", lwd=3, lty="dotted", add=T)
  qqnorm(model[["residuals"]], main = "Residual Q-Q plot")
  qqline(model[["residuals"]],col="red", lwd=3, lty="dotted")
  par(mfrow = c(1, 1))
}
```


## Model 1

We start with building linear model and include all of the variables. From the residual plots we don't see any pattern in the distribution, resiual histogram is almost normally distributed and QQplot follows almost the normal line with slight skewing at the left.

```{r}
model1_math <- lm(G3 ~., data = maths_train)
summary(model1_math)
confint(model1_math)
residualPlots(summary(model1_math))

model1_port <- lm(G3 ~., data = port_train)
summary(model1_port)
confint(model1_port)
residualPlots(summary(model1_port))

```

## Model 2

We create a second model again using the linear model and this time we choose only the significant variables using the stepAIC function.
Second model has performed better then the model 1, Residual standard error is smaller, Adjusted R-squared & F-statistic have improved. p-value is almost the same. Resiudal plots looks similar to the model 1 and since we still see the skewness we will use the glm for our next model.

```{r}

m2_math <- lm(G3~., maths_train)
model2_math <- stepAIC(m2_math, direction = "both", trace = FALSE)
summary(model2_math)
confint(model2_math)
residualPlots(summary(model2_math))

m2_port <- lm(G3~., port_train)
model2_port <- stepAIC(m2_port, direction = "both", trace = FALSE)
summary(model2_port)
confint(model2_port)
residualPlots(summary(model2_port))

```

## Model 3

We create the third model by using the generalized linear model and use the stepAIC to choose the significant variables.

```{r}
glm_math <- glm(G3~., maths_train, family = gaussian)
model3_math <- stepAIC(glm_math, direction = "both", trace = FALSE)
summary(model3_math)


glm_port <- glm(G3~., port_train, family = gaussian)
model3_port <- stepAIC(glm_port, direction = "both", trace = FALSE)
summary(model3_port)

```


# Model selection

We see similar model values for Mathematics and Portuguese datasets. So we will choose one of the model for both of the datasets.
We neglect the model1 since it as lower Adjusted R-squared, higher Residual standard error and higher AIC when compared to Model2 & Model3.

Model3 the generalized linear model is our choice here for both Mathematics and Portuguese datasets since it has a little higher Adjusted R-squared value when compared to Model2. 

Most significant variables that decide a students final grade are : **Student’s age, extra-curricular activities, quality of family relationships, weekend alcohol consumption, number of school absences, first period grade, second period grade** 



**Mathematics**

```{r message=FALSE, warning=FALSE, error=FALSE}
metrics1 <- c("Residual standard error" = summary(model1_math)$sigma,  "Adjusted R-squared" = summary(model1_math)$adj.r.squared, Predictors= length(coefficients(model1_math)), AIC=AIC(model1_math))

metrics2 <- c("Residual standard error" = summary(model2_math)$sigma,  "Adjusted R-squared" = summary(model2_math)$adj.r.squared, Predictors= length(coefficients(model2_math)), AIC=AIC(model2_math))

metrics3 <- c("Residual standard error" = sigma(model3_math),  "Adjusted R-squared" = pR2(model3_math)['r2ML'], Predictors= length(coefficients(model3_math)), AIC=model3_math$aic)


kable(cbind(metrics1, metrics2, metrics3), col.names = c("LM 1", "LM 2", "GML"))  %>% 
  kable_styling(full_width = T)
```

**Portuguese**

```{r message=FALSE, warning=FALSE, error=FALSE}
metrics1 <- c("Residual standard error" = summary(model1_port)$sigma,  "Adjusted R-squared" = summary(model1_port)$adj.r.squared, Predictors= length(coefficients(model1_port)), AIC=AIC(model1_port))

metrics2 <- c("Residual standard error" = summary(model2_port)$sigma,  "Adjusted R-squared" = summary(model2_port)$adj.r.squared, Predictors= length(coefficients(model2_port)), AIC=AIC(model2_port))

metrics3 <- c("Residual standard error" = sigma(model3_port),  "Adjusted R-squared" = pR2(model3_port)['r2ML'], Predictors= length(coefficients(model3_port)), AIC=model3_port$aic)


kable(cbind(metrics1, metrics2, metrics3), col.names = c("LM 1", "LM 2", "GML"))  %>% 
  kable_styling(full_width = T)
```
# References

P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7.

https://rdrr.io/r/stats/sigma.html

https://www.statology.org/glm-r-squared/

https://archive.ics.uci.edu/ml/datasets/student+performance


# Appendix

https://rpubs.com/petferns/DATA621-Final-Project


